Supplementary information

### 2020-03-04 manuscript/manuscript.Rmd

The decline indexes presented here provide single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
These two indexes can be represented as a two-dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease. 
Further work is needed that includes phenotypic data with greater spatial and temporal distribution to establish definitive thresholds.

@@

Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
This is to be expected and are an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007]. 
However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
It has been shown here that in the given visual definitions of oak decline severity, firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).

The high contribution of bleed prevalence related descriptors to $M_{DAI}$ is also expected as this is one of the defining features of AOD [@denman2014]. 
There was no contribution of *A. biguttatus* exit hole occurrence to the DAI and therefore stem bleeding. 
This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previously that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A. biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].

@@

A machine learning approach was used to derive the phenotypic index measures as it allows the models to be moulded to the complex interactions and trends between phenotypic descriptors from which the decline indexes are calculated. 
This would likely be oversimplified using conventional modelling strategies [@breiman2001b].

An unsupervised approach was favoured in this context due to the concession that there are known difficulties with and inaccuracy in manual decline status classification in the field [@ferretti1997].
This is highlighted in Figure 1 where numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g. COD trees with DAI values > 0.5).
The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
In situations where descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driven flexibility of the unsupervised approach.

@@

Scaling this technique to enable computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the full descriptive diversity of oak health in the population as whole.
Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
These indexes could have great utility in long term forest health monitoring by providing consistent scores which can be compared over time to inform both forestry management and research [@denman2010].

@@

```{r check,comment='',cache=FALSE}
# manuscript_check(checks = c('word count','spelling'))
```

>>>>

The decline indexes presented here provide single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
These two indexes can be represented as a two-dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease. 
Further work is needed that includes phenotypic data with greater spatial and temporal distribution to establish these definitive thresholds.

@@

Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
This is to be expected and are indicators of model suitability using the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007]. 
However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).

The high contribution of bleed prevalence related descriptors to $M_{DAI}$ is also expected as this is one of the defining features of AOD [@denman2014]. 
There was no contribution of *A. biguttatus* exit hole occurrence to the DAI and therefore stem bleeding. 
This could partly be due to the low number of trees on which *A. biguttatus* exit holes were detected (`r which(deframe(pheno_data_with_additional_descriptors[,'Agrilus exit holes']) > 0) %>% length()` out `r nrow(pheno_data_with_additional_descriptors)`) but it has also been shown previously that trees with bleeds do not always have exit holes, attributed to the fact that not all *A. biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].

@@

A machine learning approach was used to derive the phenotypic index measures as it allows the models to be moulded to the complex interactions that exist between the phenotypic descriptors from which the decline indexes are calculated. 
This would likely be oversimplified using conventional modelling strategies [@breiman2001b].

An unsupervised approach was favoured in this context due to the concession that there are known difficulties with and inaccuracy in manual decline status classification in the field [@ferretti1997].
This is highlighted in Figure 1 where numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g. COD trees with DAI values > 0.5).
The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
In situations where descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification while still allowing the data driven flexibility of the unsupervised approach.

@@

Scaling this technique to enable computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the full descriptive diversity of oak health in the population as whole.
Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
These indexes could have great utility in long term forest health monitoring by providing consistent scores which can be compared over time to inform both forestry management and research [@denman2010].
For instance, the increase of the PDI values of trees within a woodland, beyond what is considered the healthy region of the PDI scale, could indicate that mangement intervention is required.

@@

```{r check,comment='',cache=FALSE}
manuscript_check(checks = c('word count')) # ,'spelling'
```

### 2020-03-04 manuscript/supplementary.Rmd

>>>>

```{r site_differences_corrected,fig.cap='Differences between sites in phenotypic descriptors after adjustment of tree size descriptors analysed by supervised random forest classification. a) Multidimensional scaling (MDS) scatter plot of random forest proximity values of individual trees scaled to two dimensions. b) Descriptor mean decrease in accuracy importances measures. Descriptors ordered in descending importance.',fig.height=6}
loadd(site_differences_post_correction_mds_plot)
grid.arrange(site_differences_post_correction_mds_plot$a,site_differences_post_correction_mds_plot$b,nrow = 1)
```
